Boosting Higgs Pair Production in the $b\bar{b}b\bar{b}$ Channel at the LHC with Multivariate Techniques

### 2015-07-14 paper/sec-analysis.tex
small additions to bibliography
more work on paper

  These are $b$-tagged as follows.
  %
  If a small-$R$ jet has at least one $b$-quark among their constituents,
  it will be tagged as a $b$-jet with probability $f_b$.
  %
  If no $b$-quarks are found among the constituents
  of this jet, it can be still be tagged as a $b$-jet with
  a mistag rate of $f_l$.
  %
  The probability of tagging a jet is not modified
  if more than one $b$-quark is found among the jet constituents.

@@

    Large-$R$ jets are $b$-tagged by
    ghost-associating anti-$k_T$ $R=0.3$
    subjets to the original large-$R$
    jets~\cite{Cacciari:2007fd,Aad:2013gja,
      ATLAS-CONF-2014-004}.\footnote{The same
    method was used in the recent ATLAS $hh\to 4b$ search~\cite{Aad:2015uka}.}
    %
    In particular,
    a large-$R$ jet is considered  double-$b$-tagged if both
    the leading and subleading subjets (where the ordering
    is done in the subjet $p_T$) are both individually $b$-tagged.
    %
    Note that we only attempt to $b$-tag the two leading subjets:
    trying to $b$-tag all subjets above the $p_T$ cut degrades
    the signal significance due to the increase in combinatorics.

@@

Concerning the actual values of the $b$-tag probability $f_b$ and
the $b$-mistag probability of light jets $f_l$, 
we adopt as baseline the values $f_b=0.8$ and $f_l=0.01$.
%
In Sect.~\ref{sec:optimisation} we study the dependence of
our results against variations of $f_b$ and $f_l$ both
using more conservative and more aggressive values.
%

>>>>

  These are $b$-tagged as follows.
  %
  If a small-$R$ jet has at least one $b$-quark among their constituents,
  it will be tagged as a $b$-jet with probability $f_b$.
  %
  In order to be considered in the $b$-tagging algorithm,
   $b$-quarks should satisfy $p_T \ge 15$ GeV.
  %
  If no $b$-quarks are found among the constituents
  of this jet, it can be still be tagged as a $b$-jet with
  a mistag rate of $f_l$, unless a charm quark is found,
  when the mistag rate is $f_c$.
  %
  Only jets which contain at least one constituent
  with $p_T \ge 15$ GeV can lead to fake $b$-tags.
  %
  The probability of tagging a jet is not modified
  if more than one $b$-quark is found among the jet constituents.

@@

    Large-$R$ jets are $b$-tagged by
    ghost-associating anti-$k_T$ $R=0.3$
    subjets to the original large-$R$
    jets~\cite{Cacciari:2007fd,Aad:2013gja,
      ATLAS-CONF-2014-004}.\footnote{The same
    method was used in the recent ATLAS $hh\to 4b$ search~\cite{Aad:2015uka}.}
    %
    In particular,
    a large-$R$ jet is considered  double-$b$-tagged if both
    the leading and subleading subjets (where the ordering
    is done in the subjet $p_T$) are both individually $b$-tagged.
    %
    As in the case
    of small-$R$ jets, in
    order to be considered in the $b$-tagging algorithm,
   $b$-quarks should satisfy $p_T \ge 15$ GeV.
    %
    Note that we only attempt to $b$-tag the two leading subjets:
    trying to $b$-tag all subjets above the $p_T$ cut degrades
    the signal significance due to the increase in combinatorics.

@@

Concerning the actual values of the $b$-tag probability $f_b$ and
the $b$-mistag probability of light and charm jets $f_l$ and  $f_c$, 
we adopt as baseline the values $f_b=0.8$, $f_l=0.01$
and  $f_c=0.1$.
%
In Sect.~\ref{sec:optimisation} we study the dependence of
our results against variations of $f_b$ and $f_l$ both
using more conservative and more aggressive values.
%

### 2015-07-14 paper/sec-mva.tex
more work on paper

It is possible to verify that the usefulness of the MVA when
applied to the boosted category arises from the 
the information contained on the various jet substructure variables.
%
From the distribution of total associated ANN weights in the
boosted category, shown in Fig.~\ref{fig:nnweights}, we see that
all substructure variables contain useful information:
the $k_t$ splitting scales $\sqrt{d_{12}}$,
the $\tau_{21}$ subjettiness ratio, and the ECF ratios
$C_2^{(\beta)}$ and $D_2^{(\beta)}$.
%
Interestingly, for some substructure variables, such as $\tau_{21}$,
the total associated weight turns out to be stronger in the leading
than in the sub-leading Higgs candidate (or vice-versa), indicating some
degree of redundancy when the same variable is used for the two Higgs candidates
at the same time.

Indeed, the information contained Fig.~\ref{fig:nnweights} should be
reflected in the corresponding distributions of these substructure
variables: those with a higher discrimination power should show
clear differences in shape between the signal and background events,
while those that carry less information should have similar
shapes for signal and background.
%
The third possibility is that of of redundant variables, where signal
and background exhibit different shapes but the same information
is already provided by other input variables.
%
To illustrate these points,
In Fig.~\ref{fig:mva_substructure_1}
we show the distributions of some of
these substructure variables for the boosted category: the
$k_t$ splitting scale $\sqrt{d_{12}}$, Eq.~(\ref{eq:ktsplitting}) for
the subleading Higgs,
the energy correlation double ratio $D_2^{(\beta)}$
Eq.~(\ref{eq:d2}) for the leading Higgs,
and then 
the subjettiness ratio $\tau_{12}$
Eq.~(\ref{eq:tau21})
for both the leading and subleading Higgs candidates.
%

@@

From Fig.~\ref{fig:mva_substructure_1}
we can observe that for all variables the shapes of the signal
and background distributions are quite different,
reflecting the inherent differences in the internal structure of jets
between QCD jets and jets originated from the Higgs decays.
%
Note that as before, all the distributions in
Fig.~\ref{fig:mva_substructure_1}
have been normalized to unity.
%
Importantly, we observe that signal and background distributions peak
in different regions: for example, the $k_t$ splitting scale $\sqrt{d_{12}}$
peaks around 80 GeV (50 GeV) for signal (background) events, while
the ECF ratio $D_2^{(\beta)}$ peaks at much lower values for signal than
for background events.
%
From Fig.~\ref{fig:mva_substructure_1} we also see
the distributions of the subjettiness ratio $\tau_{12}$ are quite similar
in the leading and in the subleading jets, therefore adding the value
of  $\tau_{12}$ from the two Higgs candidates at the MVA at the same time
is redundant, explaining why in these cases either one or the other
variable from a Higgs candidate is assigned higher weight.

Finally, let us show the distributions of kinematic variables
that according to the MVA provide a small amount of discrimination
between signal and background, and where the should not be an issue
of redundancy.
%
In Fig.~\ref{fig:mva_substructure_2}
we show the kinematic variable that carries less weight in the MVA,
    the azimuthal angle separation between the two Higgs candidates
    $\Delta\phi_{hh}$, for the resolved and boosted categories:
    according to Fig.~\ref{fig:nnweights}, no much discrimination power
    is obtained here, and indeed the shapes of the two signal
    and background
    are very similar here.
    %
    From Fig.~\ref{fig:nnweights} we also saw that the
    invariant mass distributions $m_h$ carry little information,
    and this was verified from 
the corresponding
distributions in $m_{hh}$ from Fig.~\ref{fig:mHHinv}.
%
As explained before, the reason for this is
the  smearing of jet four-momenta introduced to mimic
realistically detector resolution.

>>>>

%
From the distribution of total associated ANN weights in the
boosted category, shown in Fig.~\ref{fig:nnweights}, we see that
all substructure variables contain useful information:
the $k_t$-splitting scales $\sqrt{d_{12}}$,
the $\tau_{21}$ subjettiness ratio, and the ECF ratios
$C_2^{(\beta)}$ and $D_2^{(\beta)}$.
%
Interestingly, for some substructure variables, such as $\tau_{21}$,
the total associated weight turns out to be stronger in the leading
than in the sub-leading Higgs candidate (or vice-versa), indicating some
degree of redundancy when the same variable is used for the two Higgs candidates
at the same time.

The information reflected in Fig.~\ref{fig:nnweights} should also be
apparent from the corresponding distributions of these substructure
variables: those with a higher discrimination power should show
clear differences in shape between the signal and background events,
while those that carry less information should have similar
shapes for signal and background.
%
The third possibility is that of of redundant variables, where signal
and background exhibit different shapes but the same information
is already provided by other input variables.
%
To illustrate these points,
in Fig.~\ref{fig:mva_substructure_1}
we show the distributions of some of
these substructure variables for the boosted category: the
$k_t$ splitting scale $\sqrt{d_{12}}$, Eq.~(\ref{eq:ktsplitting}), for
the subleading Higgs,
the energy correlation double ratio $D_2^{(\beta)}$,
Eq.~(\ref{eq:d2}), for the leading Higgs,
and then 
the subjettiness ratio $\tau_{12}$,
Eq.~(\ref{eq:tau21}),
for both the leading and subleading Higgs candidates.
%

@@

From Fig.~\ref{fig:mva_substructure_1}
we can observe that for these substructure variables the shapes of the signal
and background distributions show important differences,
reflecting the inherent differences in the internal structure of jets
between QCD jets and jets originated from the Higgs decays.
%
Importantly, we observe that signal and background distributions peak
in different regions: for example, the $k_t$ splitting scale $\sqrt{d_{12}}$
peaks around 80 GeV (50 GeV) for signal (background) events, while
the ECF ratio $D_2^{(\beta)}$ peaks at much lower values for signal than
for background events.
%
From Fig.~\ref{fig:mva_substructure_1} we also see
the distributions of the subjettiness ratio $\tau_{12}$ are quite similar
in the leading and in the subleading jets, therefore adding
$\tau_{12}$ from the two Higgs candidates to the MVA at the same time
is redundant, explaining why in these cases either one or the other
variable from a Higgs candidate is assigned higher weight.

It is also interesting to show the distributions of kinematic variables
that according to the MVA provide a small amount of discrimination
between signal and background, and where the should not be an issue
of redundancy.
%
In Fig.~\ref{fig:mva_substructure_2}
we show the kinematic variable that carries less weight in the MVA,
    the azimuthal angle separation between the two Higgs candidates
    $\Delta\phi_{hh}$, for the resolved and boosted categories:
    according to Fig.~\ref{fig:nnweights}, essentially no discrimination power
    should be obtained from this variable, and indeed the shapes of the two signal
    and background
    are very similar here.
    %
    From Fig.~\ref{fig:nnweights} we also saw that the
    invariant mass distributions $m_h$ carry little information
    to discriminate signal over background:
    this was already verified from 
the corresponding
distributions in Fig.~\ref{fig:mHHinv}.
%
As explained before, the reduced discrimination power arises
from 
the  smearing of jet four-momenta introduced to mimic
realistically detector resolution.\footnote{We have verified that
  if no smearing is applied, the $m_h$ distributions for the leading
  and subleading Higgs candidates become the most
important variables in the MVA.}

### 2015-07-14 paper/sec-optimisation.tex
more work on paper

\section{Optimisation}
\label{sec:optimisation}

One important motivation of this work was to quantify which aspects
of the performance of the LHC detectors needs to be improved
the most in order to maximise the significance of the observation
of double Higgs production in the $4b$ final state.
%
In this section we explore how the results with the baseline settings,
summarized in the previous section, are modified when some of
these settings are changed.
%
In particular we will explore the dependence of the results on
the $b$-tagging probability $f_b$, the light jet mistag rate
$f_l$, as well on the momentum smearing fraction.
%
The variations of the analysis
settings that we will explore are collected in
Table~\ref{sec:variations}.

@@

We also study how the signal significance is modified as a function
of the systematic uncertainty that needs to be obtained
in the measurement of the background processes.
%
This is specially important in a process like this one where
the signal over background ratio $S/B$ is small, and thus were
it is essential to measure the backgrounds as accurately
as possible.

Given that we have found in the previous section that the
only category with real discrimination power is the boosted one,
in this section we concentrate only on this one,
by comparing our baseline results from Table~\ref{table:cutflowMVA}
with the new scenarios.
%
When necessary, the MVA has been retrained to optimise the
new information contained in the various scenarios.

The results of this optimisation study are collected in
Table~\ref{table:cutflowMVAoptimisation}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{table}[t]
  \centering
  \begin{tabular}{c||c|c|c|c}
    \hline
    \multicolumn{5}{c}{Boosted category}\\
    \hline
    \hline
 Scenario &    \multicolumn{2}{c|}{$N_{\rm ev}$} &  $S/\sqrt{B}$  & $S/B$ \\
       &   Signal & Back   &     &    \\
 \hline
 \hline
   Baseline   & 107 & 1040 & 3.31  & 0.10\\
   \hline
   &  &   &   &   \\
   &  &   &   &   \\
   &  &   &   &   \\
   &  &   &   &   \\
         &  &   &   &   \\
   \hline
  \end{tabular}
  \caption{\small The details of the various scenarios have been collected in
    Table~\ref{sec:variations}.
    \label{table:cutflowMVAoptimisation}
  }
\end{table}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

>>>>

\section{Analysis optimisation}
\label{sec:optimisation}

One motivation for this work was to quantify which aspects
of the performance of the LHC detectors needs to be improved
the most in order to maximise the significance of the observation
of double Higgs production in the $4b$ final state.
%
In this section we explore how the post-MVA
results with the baseline settings,
summarized in the previous section, are modified when some of
these settings are changed.
%
In particular we will explore the dependence of the results on
the $b$-tagging probability $f_b$, the light jet mistag rate
$f_l$, as well on the four-momentum smearing fraction $\Delta p$.
%
The variations of the analysis
settings that we will explore are collected in
Table~\ref{sec:variations}.

@@

Given that we have found in the previous section that the
only category with real discrimination power is the boosted one,
in this section we concentrate only on this one,
by comparing our baseline results from Table~\ref{table:cutflowMVA}
with the new scenarios.
%
When necessary, the MVA has been retrained to optimise the
new information contained in the various scenarios.
%
This is important since for example, when we vary $f_b$
or $f_l$, the relative composition of the multijet
background will change, and this can translate into
modifications of the kinematical variables used for
the MVA training.
%
Similarly, if the momentum resolution is modified,
the weight of the different variables in the MVA
will be altered, in particular the invariant mass
of Higgs candidates $m_h$.

The results of this optimisation study are collected in
Table~\ref{table:cutflowMVAoptimisation}.
%
There we compare the results for the number of signal
and background events $N_{\rm ev}$ expected at the
HL-LHC, the signal significance and the $S/B$ ratio
for our nominal settings and for the variations
of the analysis settings.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{table}[t]
  \centering
  \begin{tabular}{c||c|c|c|c}
    \hline
    \multicolumn{5}{c}{Boosted category}\\
    \hline
    \hline
 Scenario &    \multicolumn{2}{c|}{$N_{\rm ev}$} &  $S/\sqrt{B}$  & $S/B$ \\
       &   Signal & Back   &     &    \\
 \hline
 \hline
   Baseline   & 107 & 1040 & 3.31  & 0.10\\
   \hline
   &  &   &   &   \\
   &  &   &   &   \\
   &  &   &   &   \\
   &  &   &   &   \\
         &  &   &   &   \\
   \hline
  \end{tabular}
  \caption{\small
Number of signal
and background events $N_{\rm ev}$ expected at the
HL-LHC, signal significance and $S/B$ ratio
for the nominal settings and for the variations
of the analysis settings, summarized
in  Table~\ref{sec:variations}.
 \label{table:cutflowMVAoptimisation}
  }
\end{table}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Neglecting the
contribution from light jet fakes,
it is easy to estimate how $S/\sqrt{B}$ should
scale with $f_b$: since we have two double-$b$-tagged jets
both in signal and background events, one should find that
\be
\frac{S}{\sqrt{B}}\Bigg|_{f_b}\Bigg/
\frac{S}{\sqrt{B}}\Bigg|_{f_b'}
\simeq \frac{f_b^2}{f_b'^2} \, .
\ee
By comparing this expectations with the numbers of
Table~\ref{table:cutflowMVAoptimisation}, we see that there
is a reasonable agreement in the boosted category,
as expected since it is dominated by the irreducible
$4b$ component.
